### 2025-04-07 Thesis/Latex/Chapter_2.tex
update

\chapter{Large Language Models (LLMs)}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{start1}
Large Language Models (LLMs) represent a transformative leap in Natural Language Processing (NLP), allowing machines to perform complex language tasks with remarkable accuracy. From generating coherent text to answering nuanced questions, LLMs have pushed the boundaries of what AI can achieve. This chapter delves into the architecture and principles behind LLMs, their development through scaling and training techniques, and the models that have defined this field. It also addresses the challenges these models face and their expanding role in various real-world applications.

\section{Natural language processing (NLP)}
Natural Language Processing (NLP) is a field within artificial intelligence (AI) that concentrates on the interaction between computers and human language. It involves the development of algorithms and models  that allow machines to comprehend, interpret, and generate human language in a meaningful context.NLP is crucial for enabling computers to process and respond to human language effectively, as demonstrated by features like Google's predictive text in keyboards and language translation systems that manage multiple languages efficiently \cite{murugan2024nlp}

\subsection{Historical development of NLP}
The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s. It then highlights the impact of deep learning from the 2000s onward, leading to the development of pre-trained models like BERT and GPT. The timeline concludes with the emergence of large language models (LLMs) from 2019 to the present, marking a significant shift in NLP research and applications.
\begin{figure}[htbp]
	\centerline{\includegraphics[scale=.7]{Figures/the_historyOf_nlp.png}}
	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
	\label{the_historyOf_nlp.png}
\end{figure}

\subsection{Importance of natural language processing}

\begin{enumerate}
	\item \textbf{Efficient Data Processing:} NLP enables businesses to process and analyze large volumes of unstructured, text-heavy data that would be challenging to handle otherwise..

	\item \textbf{Understanding Human Language:}NLP can interpret complex language elements, such as acronyms, abbreviations, and context, improving the accuracy of machine learning models..

	\item \textbf{Advancements in Technology:}Improvements in deep learning and machine learning have made NLP more effective, expanding the types of data that can be analyzed.

	\item \textbf{Natural Interactions:} NLP allows users to interact naturally with AI chatbots and voice assistants, like Siri, without needing to use specific,predefined\cite{techtarget_nlp}.

\end{enumerate}

\section{Neural Networks in NLP}
A neural network, or artificial neural network, is a machine learning algorithm inspired by the human brain. It is a key component of deep learning, a branch of machine learning effective in solving complex problems like image recognition and language processing. Unlike traditional computer programs that use a step-by-step algorithmic approach, neural networks learn from examples, mimicking the way neurons in the human brain operate. They consist of interconnected nodes (processing elements) that work together in parallel to solve specific problems. \\
A neural network has three basic sections, or parts, each composed of "nodes." \\ The input layer is the first part, receiving raw data, with each node (or neuron) representing a feature of this data. The hidden layers, which form the intermediate section, perform various transformations and computations, enabling the network to learn complex patterns and relationships. Finally, the output layer, which is the last part, produces the network's output, with the number of nodes corresponding to the desired output classes or regression values \cite{murugan2024nlp}.
\begin{figure}[htbp]

	\centerline{\includegraphics[width=.7\linewidth]{Figures/ANN.png}}
	\caption{Artificial Neural Network (ANN).}
	\label{ANN.png}
\end{figure}

\subsection{Basics Concepts in Neural Networks}
\textbf{Weights} \\
Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.  \\ \\ 
\textbf{Biases} \\
Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer. Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning. 
\textbf{Activation Functions} \\
These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns. Common activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU). \cite{taylor2017neural}.

\subsection{ Recurrent Neural Networks (RNNs)}
Recurrent neural networks (RNNs) are a type of neural network designed for processing sequential data, such as text or time series, by effectively handling variable-length inputs. An RNN consists of a hidden state h and an optical output y, which operate on a variable  sequence x=($x_1$....$x_t$)At each time step t, the hidden state $h^t$
of the RNN is updated according to the function $h^{(t)} = f(h^{(t-1)}, x_t)$
where f is a non-linear activation function, such as a logistic sigmoid or a more complex long short-term memory (LSTM) unit. This design allows RNNs to remember past inputs and incorporate them into future outputs, enabling them to recognize patterns that occur at multiple positions within a sequence. Additionally, RNNs utilize parameter sharing across time steps, which helps in generalizing across sequences of varying lengths and learning complex dependencies over time. Despite challenges with long-term dependencies, RNNs are particularly effective for tasks requiring context or sequential understanding, making them valuable for various applications, including time series analysis and natural language processing\cite{cho2014learning} 
\begin{figure}[htbp]

	\centerline{\includegraphics[width=1\linewidth]{Figures/RNN.png}}
	\caption{RNN Architecture.}
	\label{RNN.png}
\end{figure}
\subsection{ Long-Short Term Memory (LSTM)}
Long Short-Term Memory (LSTM) networks, designed by Hochreiter and Schmidhuber, are an improved version of recurrent neural networks (RNNs) designed to learn long-term dependencies in sequential data, making them suitable for tasks like time series forecasting and language translation
They address the limitations of traditional RNNs by introducing a memory cell that maintains information over extended periods. This memory cell is regulated by three gates—input, forget, and output gates—which control the flow of information in and out of the cell as shown in the fig \cite{geeksforgeeks_lstm}
\begin{figure}[htbp]
	\centering\includegraphics[width=0.9\linewidth]{Figures/LSTM.png}
	\caption{LSTM Architecture.}
	\label{/LSTM.png}
\end{figure}
Forget Gate: Decides what information to discard from the cell state
\[
f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
\]

Input Gate: Determines what new information to add to the cell state
\[
i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
\]
\[
\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
\]

Output Gate: Controls what information to output from the cell state
\[
o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
\]
\[
h_t = o_t \odot \tanh(C_t)
\]

It is worth mentioning that Bidirectional LSTM networks extend the traditional LSTM architecture by processing data in both forward and backward directions. This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies. Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}. 
\section{ Transformer Architecture}
The Transformer architecture is a deep learning model introduced in June 2017 by Vaswani et al. from Google Brain. Their paper, titled "Attention Is All You Need," presented a groundbreaking approach to processing sequential data through the use of a self-attention mechanism. This innovative method allows the model to assign different levels of importance to various parts of the input, enabling it to capture long-range dependencies much more effectively than earlier models like RNNs and LSTMs.The original Transformer model is structured as a stack of six layers, where the output of each layer i serves as the input to the subsequent layer i+1, continuing this process until the final prediction is reached.It features a six-layer encoder on the left and a corresponding six-layer decoder on the right,  both of which work together to transform input sequences into meaningful outputs. Each encoder and decoder consists of six identical layers that allow the model to process and generate language efficiently \cite{rothman2021transformers}.
\begin{figure}[htbp]

	\centerline{\includegraphics[width=.6\linewidth]{
			Figures/trasformers.png}}
	\caption{The Transformer - model architecture..}
	\label{trasformers.png}

\end{figure}
\subsection{Encoder and Decoder Stacks}
\textbf{Encoder:} Each layer in the encoder consists of two sub-layers: \\
Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
Position-wise feed-forward network: A fully connected network applied to each position separately and identically. \\
Residual Connections: Residual connections are applied around each sub-layer, followed by layer normalization. The output of each sub-layer is computed as Layer Norm(x + Sublayer(x)). \\
Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
\textbf{Decoder:} Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
Multi-head self-attention mechanism ,
Position-wise feed-forward network. \\
Additional Sub-layer: The decoder includes a third sub-layer that performs multi-head attention over the output of the encoder stack. \\
Residual Connections and Normalization: Like the encoder, residual connections are applied around each sub-layer, followed by layer normalization. \\
Masked Self-Attention: The self-attention mechanism in the decoder is modified to prevent positions from attending to subsequent positions, ensuring that predictions for position i depend only on the known outputs at positions before i \cite{vaswani2017attention}.
\begin{figure}[htbp]

	\centerline{\includegraphics[width=.8\linewidth]{
			Figures/incoderDecoder.png}}
	\caption{Encoder and Decoder Stacks}
	\label{incoderDecoder.png}

\end{figure}
\subsection{Self-Attention Mechanism}
Self-attention is a mechanism that allows the model to weigh the importance of different words in a sequence relative to each other. This is crucial for understanding context and relationships between words \cite{vaswani2017attention}.
\subsection{Query, Key, and Value Vectors:}
\textbf{Query Vectors (Q):} Generated from the input word's embedding and represent the word's role in finding relevant information in other words.\\
\textbf{Key Vectors (K):} Derived similarly from embedding and serve as a comparison standard for each word.\\
\textbf{Value Vectors (V):} Contain the actual information to be passed on and are weighted according to relevance determined by the Query and Key comparison \cite{vaswani2017attention}.. 
\subsection{Self-Attention Calculation:}
The score is computed by taking the dot product of the Query vector of a word with the Key vectors of all other words. This score determines how much focus should be placed on each word when encoding the current word.
The scores are scaled and passed through a softmax function, producing a distribution of attention weights. \\
Each Value vector is weighted by these attention scores and summed to produce the final output of the self-attention layer\cite{vaswani2017attention}.
\subsection{Role of Multi-Head Attention}
Multi-head attention enhances the model’s ability to focus on different aspects of the input by creating multiple sets of Query, Key, and Value vectors. Each set is processed independently and then combined. \\
This approach allows the model to attend to information from different representation subspaces at different positions, improving its ability to capture complex relationships\cite{vaswani2017attention}.
\subsection{Positional Encoding:}
Since self-attention mechanisms do not inherently capture the order of words, positional encoding is added to the word embeddings to provide information about the relative position of each word in the sequence.
These encodings are added directly to the input embeddings, allowing the model to consider both the position and the content of 
each word during processing\cite{alammar2018illustratedtransformer}.
\subsection{Transformers in NLP}
Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data. This advancement has led to the creation of highly effective models like BERT and GPT, with its deep bidirectional context, excels at understanding and improving performance across various NLP tasks. GPT, on the other hand, is renowned for generating coherent and contextually relevant text, significantly advancing applications such as text generation and translation\cite{devlin2019bert}.\\
\textbf{Scalability:} Transformers efficiently handle large datasets and long sequences, overcoming the limitations of RNNs. This scalability allows for training with billions of parameters, enhancing model capabilities.\\
\textbf{Rich Contextual Understanding: }The self-attention mechanism in transformers captures relationships between words across the entire sequence, enabling deep contextual understanding and more accurate language processing.\\
\textbf{Model Efficiency:} Transformers enable parallel processing, which speeds up training and makes them more efficient than sequential models like RNNs. This efficiency supports the rapid development and deployment of advanced language models\cite{vaswani2017attention}.
\section{ Emergence of Large Language Models (LLMs)}

Large Language Models (LLMs) are advanced artificial intelligence systems designed to process and generate human-like text. They are typically based on transformer architectures and are characterized by their enormous scale, with billions of parameters. LLMs are pre-trained on vast amounts of text data in a self-supervised manner, enabling them to develop a broad understanding of language. They are capable of performing a wide range of tasks with minimal task-specific fine-tuning, often achieving significant performance improvements through few-shot or zero-shot learning\cite{brown2020language}.
\subsection{Scaling in Large Language Models (LLMs)}
Scaling is crucial in the evolution of large language models. The history of scaling shows that increasing both model size and dataset size leads to significant improvements in performance across various NLP tasks. For instance, early work by Brants et al. (2007) demonstrated the benefits of using language models trained on vast datasets, such as 2 trillion tokens, which led to significant advancements in machine translation quality. This was followed by efforts like those of Heafield et al. (2013), who scaled traditional models to Web-scale data, and Jozefowicz et al. (2016), who scaled LSTMs to 1 billion parameters, achieving state-of-the-art results on large benchmarks.The advent of transformer-based models marked a significant shift.Models like BERT, GPT-2, and GPT-3, with their enormous parameter counts—up to 175 billion for GPT-3—demonstrated that scaling up not only the model but also the dataset size yields substantial gains in performance. Researchers like Kaplan et al. (2020) and Hoffmann et al. (2022) studied how scaling affects model performance, proposing power laws that show a predictable relationship between model size, dataset size, and performance. These studies emphasized the importance of scaling for the continued progress of LLMs\cite{touvron2023llama}.
\subsection{Pre-training}
Pre-training is a crucial stage in developing Large Language Models (LLMs), where the model learns from extensive unlabeled datasets through a process called self-supervision. This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance
\textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens. This autoregressive method enables models like GPT-3 to generate coherent and contextually relevant text.\\
\textbf{Prefix Language Modeling: }Employed in encoder-decoder and non-causal decoder-only models, this technique uses a non-causal (considering both past and future tokens) prefix for predicting subsequent tokens, offering more flexibility and enhancing the model's adaptability across various language tasks.\\
\textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context. An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.
\begin{figure}[htbp]

	\includegraphics[width=1\linewidth]{Figures/pretraining.png}
	\caption{Training Tokenization in Full, Prefix, and Masked Language Modeling.}
\end{figure}
\subsection{Fine-tuning }
Fine-tuning has been the most common approach for adapting LLMs to specific tasks. After pre-training on large datasets, these models are further trained on smaller, supervised datasets tailored to the desired task. Fine-tuning updates the weights of the pre-trained model, allowing it to excel at specific tasks like sentiment analysis, machine translation, or question answering. The main advantage of fine-tuning is that it often results in strong performance on task-specific benchmarks. However, it comes with challenges, such as the need for a new large dataset for every task, the risk of overfitting to narrow distributions, and potential poor generalization to out-of-distribution examples\cite{touvron2023llama}.
\subsection{Few-Shot, One-Shot, and Zero-Shot Learning}
Few-shot, one-shot, and zero-shot learning represent different paradigms of using LLMs without extensive fine-tuning. These methods involve using pre-trained models to perform tasks with minimal task-specific data.\\
\textbf{Few-Shot Learning} involves giving the model a few examples (typically 10-100) of the task during inference. This approach reduces the need for large, task-specific datasets and limits the risk of overfitting to narrow data distributions. However, few-shot results are generally not as strong as those from fully fine-tuned models.\\
\textbf{One-Shot Learning} is similar to few-shot learning but uses only a single example alongside a natural language description of the task. This method mirrors how humans are often instructed for tasks, making it an interesting approach for tasks where providing multiple examples is impractical.\\
\textbf{Zero-Shot Learning} requires the model to perform a task based solely on a natural language description, with no examples provided. While this method offers maximum flexibility and robustness, it is also the most challenging for models. Zero-shot performance is often weaker than few-shot or one-shot, but it represents a significant step towards task-agnostic AI, where models can generalize across a wide range of tasks with minimal human intervention\cite{brown2020language}.
\subsection{Evaluation Datasets and Tasks}
Evaluating Large Language Models (LLMs) is essential for determining their effectiveness and limitations in comprehending and generating human language. This evaluation typically falls into two primary categories:
\begin{itemize}
	\item Natural Language Understanding (NLU):This measures the model's proficiency in comprehending language, encompassing tasks such as sentiment analysis, text classification, natural language inference (NLI), question answering (QA), commonsense reasoning (CR), mathematical reasoning (MR), and reading comprehension (RC).
	\item Natural Language Generation (NLG):This assesses the model's capability to produce text based on given context. It includes tasks like summarization, sentence completion, machine translation (MT), and dialogue generation.
\end{itemize}
\textbf{Benchmarks} play a critical role in evaluating LLMs, providing standardized tests to measure their performance across various tasks:
\begin{itemize}
	\item MMLU: Measures model knowledge from pretraining and evaluates performance in zero-shot and few-shot scenarios across 57 subjects, testing world knowledge and problem-solving abilities.

>>>>

\chapter{Retrieval-Augmented Generation (RAG)}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{ Retrieval-Augmented Generation (RAG)}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{start 1}

Large language models (LLMs) have seen significant advancements but face challenges, particularly in tasks that demand extensive knowledge or deal with queries beyond their training data. These limitations often result in inaccuracies or “hallucinations.” To address this, Retrieval-Augmented Generation (RAG) supplements LLMs by retrieving relevant document chunks from external knowledge sources based on semantic similarity. By doing so, RAG helps reduce factual errors, allowing LLMs to produce more accurate content. This has led to the widespread adoption of RAG, particularly in chatbots and other real-world applications, making it a crucial technology in advancing the capabilities of LLMs.
\section{Fundamentals of Retrieval-Augmented Generation}
RAG represents a significant advancement in the capabilities of large language models (LLMs) by integrating external knowledge retrieval with the generation of text. Understanding the individual components of retrieval and generation is essential to appreciate how their synergy improves the overall performance of RAG systems.
\subsection{Definition of (RAG)}
Retrieval-Augmented Generation (RAG) is a technique that leverages the strengths of pre-trained large language models (LLMs) and external data sources. By merging the generative abilities of LLMs like GPT-3 or GPT-4 with the accuracy of specialized data search mechanisms, RAG systems can generate more sophisticated and contextually relevant responses\cite{selvaraj2024}.
\subsection{Historical Development of (RAG)}
The foundations of Retrieval-Augmented Generation (RAG) can be traced back to traditional information retrieval (IR) systems,  Early models primarily relied on keyword matching and simple ranking mechanisms to retrieve relevant documents from databases, establishing a basic framework for information access. The landscape began to shift with the rise of neural networks in the 2010s. Notable innovations like Word2Vec and Transformer-based architectures enabled retrieval methods to incorporate deeper semantic understanding, paving the way for more sophisticated document retrieval processes.
A significant advancement occurred with the introduction of dense passage retrieval (DPR) in 2020, which utilized bi-encoder architectures to map both queries and documents into dense vector spaces. 
The integration of these advanced retrieval techniques with large language models (LLMs) became a focal point as models like GPT-3 gained prominence. Researchers explored how LLMs could be augmented with retrieval capabilities, leading to the development of RAG systems that leverage the strengths of both retrieval and generative capabilities\cite{gao2024retrieval}.
\subsection{Differences between (RAG), fine-tuning and transfer learning}
Recent advances in natural language processing have led to the development of three prominent methodologies for enhancing language models: Retrieval-Augmented Generation (RAG), fine-tuning, and transfer learning.

RAG, as described by Lewis et al. \cite{lewis2020retrieval}, integrates external retrieval mechanisms with generative models. This approach enables the model to dynamically access and incorporate relevant information from external knowledge bases during the generation process, thereby addressing limitations in its internal memory and providing more accurate responses for knowledge-intensive tasks.

Fine-tuning, detailed by Howard and Ruder \cite{howard2018universal}, involves adapting a pre-trained model to a specific task by further training it on a targeted dataset. This process refines the model’s parameters to better capture the nuances of the task at hand, leading to improved performance even when the available domain-specific data is relatively limited.

Transfer learning, as exemplified by Devlin et al. \cite{devlin2018bert}, refers to the broader strategy of leveraging a model that has been pre-trained on large, diverse datasets for use in various downstream tasks. This method capitalizes on the rich, generalized representations learned during the pre-training phase, and then applies task-specific adaptations to achieve state-of-the-art results across a range of applications.

Collectively, these methodologies offer complementary strategies to enhance language model performance, each contributing unique advantages and addressing different challenges in the field of natural language processing.

\section{Types of (RAG) Systems}
As detailed in \cite{gao2024retrieval}, the evolution of Retrieval-Augmented Generation (RAG) can be categorized into three distinct phases: Naive RAG, Advanced RAG, and Modular RAG .
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/rag_types.png}
	\caption{Types of RAG Systems}
	\label{rag_architectur.png}
\end{figure}
\subsection{Naive (RAG)}
Naive RAG, a foundational approach in Retrieval-Augmented Generation, operates on a straightforward "Retrieve-Read-Generate" paradigm. This method involves three primary steps:

\begin{itemize}
	\item Indexing: Raw data is cleaned, extracted, and converted into a uniform text format. It's then segmented into smaller chunks and encoded into vector representations using an embedding model. These vectors are stored in a vector database for efficient similarity searches.
	\item Retrieval: When a user query is received, it's encoded into a vector and compared to the indexed chunks. The top K most similar chunks are retrieved and included in the prompt for the language model.
	\item Generation: The query and retrieved chunks are combined into a prompt, which is fed to a large language model. The model generates a response based on the provided context and its internal knowledge.
\end{itemize}
While Naive RAG offers a basic framework, it faces several challenges in 
\begin{itemize}
	\item Retrieval Issues: The retrieval process can be imprecise, leading to the selection of irrelevant or missing information.
	\item Generation Challenges: The language model may hallucinate, generating content not supported by the retrieved context. It may also produce irrelevant, toxic, or biased outputs.
	\item Augmentation Difficulties: Integrating retrieved information into the generation process can be challenging, leading to disjointed or redundant responses.
\end{itemize}

\subsection{Advanced (RAG)}
Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.\\
\textbf{Pre-retrieval Strategies:}
\begin{itemize}
	\item Enhanced Indexing: Advanced RAG tackles indexing issues through a sliding window approach, finer segmentation of data, and inclusion of metadata. Additionally, it optimizes the retrieval process by employing various methods.
	\item Query Optimization: This stage focuses on refining the user's initial query to make it clearer and more suitable for retrieval. Techniques like query rewriting, transformation, and expansion are commonly used.
\end{itemize}
\textbf{Post-Retrieval Strategies:}
\begin{itemize}
	\item Re-ranking Chunks: After relevant information is retrieved, Advanced RAG prioritizes the most relevant content by re-ranking the retrieved chunks and placing them strategically within the prompt.
	\item Context Compression: To avoid overwhelming the LLM with too much information, post-retrieval efforts focus on selecting the most essential parts of the retrieved context, highlighting critical sections, and compressing the data to be processed.
\end{itemize}
By addressing indexing issues and refining the query and retrieved information, Advanced RAG aims to improve the overall accuracy and relevance of the generated response.
\subsection{Modular (RAG)} 
Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.\\
\textbf{New Modules}
\begin{itemize}
	\item Search Module: Adapts to specific scenarios by leveraging LLM-generated code and query languages to search across various data sources.
	\item RAG Fusion: Employs a multi-query strategy to expand user queries, uncover both explicit and implicit knowledge, and improve retrieval results.
	\item Memory Module: Leverages the LLM's memory to guide retrieval and create an unbounded memory pool, aligning the text more closely with data distribution.
	\item Routing Module: Navigates through diverse data sources, selecting the optimal pathway for a query based on its specific needs.
	\item Predict Module: Reduces redundancy and noise by generating relevant context directly through the LLM.
	\item Task Adapter Module: Tailors RAG to various downstream tasks by automating prompt retrieval and creating task-specific retrievers.
\end{itemize}
\textbf{New Patterns} 
\begin{itemize}
	\item Flexible Module Arrangement: Modular RAG allows for the substitution and reconfiguration of modules to address specific challenges, surpassing the fixed structures of previous RAG paradigms.
	\item Innovative Retrieval Strategies: Techniques like Rewrite-Retrieve-Read, Generate-Read, and Recite-Read leverage the LLM's capabilities to refine queries, generate content, and retrieve information from model weights.
	\item Hybrid Retrieval: Combines keyword, semantic, and vector searches to cater to diverse queries, improving retrieval relevance.
	\item Dynamic Module Interaction: Frameworks like Demonstrate-Search-Predict and ITERRETGEN demonstrate the dynamic use of module outputs to enhance each other's functionality.
	\item Adaptive Retrieval: Techniques like FLARE and Self-RAG evaluate the necessity of retrieval based on different scenarios, allowing for a more flexible and efficient approach.

\end{itemize}
\section{Core Components of RAG}
The core components of Retrieval-Augmented Generation (RAG) systems consist of a retrieval mechanism, a generation process, and augmentation techniques. These elements work together to enhance the model's ability to access relevant external information, generate coherent and contextually appropriate responses, and improve overall performance in knowledge-intensive tasks.
\subsection{ Retrieval Mechanism}
RAG systems combine parametric memory (a pre-trained language model) with non-parametric memory (a retrieval mechanism). The retrieval mechanism allows RAG models to access external information sources (e.g., Wikipedia), and this process is central to improving the model's ability to generate factual and accurate outputs \cite{lewis2020retrieval}.
\begin{itemize}
	\item Traditional Techniques: \\
	Classical retrieval methods include \textbf{TF-IDF and BM25}, which rely on sparse vector representations based on term frequencies. These methods use exact keyword matching, making them limited in semantic understanding.
	\item Modern Retrieval Approaches:\\
	\textbf{ Dense Passage Retrieval (DPR): }This approach utilizes dense vector representations learned by neural models like BERT to encode both the query and document. It allows for a more semantic understanding of the content, making retrieval more effective. DPR, for example, computes the similarity between the query and documents using Maximum Inner Product Search (MIPS), which finds the closest passages based on the dense vector space.
	\item Implementation in RAG:\\ In RAG, dense retrieval methods are often paired with techniques like \textbf{Maximum Inner Product Search} (MIPS), which efficiently matches query and document embeddings. This enables the system to return the most relevant documents for subsequent generation \cite{karpukhin2020dense}.

\end{itemize}
\subsection{Generation Process}
The generation in RAG occurs through a combination of retrieved passages and the input query.\\
Large Language Models (LLMs) like BART or T5 are utilized for this generation, leveraging their advanced capabilities in understanding and producing human-like text. The retrieval component supplies external context, which the generator conditions on to formulate a response.This context is crucial, as it enriches the LLM's understanding, enabling it to incorporate real-time, relevant information. Moreover, the generation process benefits from the LLM's ability to synthesize information, allowing it to create responses that are not only coherent but also informed by the latest data, thus improving accuracy and relevance in knowledge-intensive tasks \cite{lewis2020retrieval}. \\
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
	\caption{Rag architecture}
	\label{rag_architectur.png}
\end{figure}
\subsection{ Augmentation Techniques}
The augmentation of the retrieval process in Retrieval-Augmented Generation (RAG) systems focuses on improving how queries are refined and how relevant information is retrieved for downstream generation tasks. Key augmentation methods include.
\begin{itemize}
	\item \textbf{Query Augmentation:} In traditional retrieval pipelines, user queries are often under-specified or ambiguous, leading to poor retrieval performance. Query augmentation involves dynamically rewriting the user's query to better match the documents in the knowledge base. This can be done by leveraging large language models (LLMs) to generate tailored queries or synthetic questions and answers (QAs) that better align with the search objective​
	\item \textbf{Synthetic QA Generation:} Instead of using raw document chunks, retrieval is enhanced by generating and embedding synthetic QA pairs from documents. This helps to capture the semantic essence of long texts more effectively, reducing noise and improving retrieval precision. These synthetic QAs can be used to rewrite the user query, making it more specific to the task at hand \cite{mombaerts2024meta}.
\end{itemize}
\section{RAG for Legal Information Retrieval}

Retrieval-Augmented Generation (RAG) holds significant promise for transforming the way legal information is accessed, processed, and utilized. In legal contexts, particularly where precision and comprehensiveness are critical, RAG enables the dynamic integration of relevant external knowledge into language models during inference. This is especially valuable for retrieving and synthesizing content from case law, legislative texts, legal commentaries, and scholarly analyses\cite{lexemoRAG}.

Unlike traditional keyword-based or Boolean legal search systems, which rely on exact matches and rigid query structures, RAG models operate by embedding both queries and documents into a shared vector space, allowing for more semantic and contextual retrieval. This enables users to pose natural language queries and obtain responses grounded in specific, retrieved legal documents. For instance, rather than retrieving a list of documents containing the keyword "contract termination," a RAG model can retrieve multiple relevant legal provisions, judgments, or commentaries and synthesize a coherent and accurate summary or answer.

Moreover, RAG is well-suited to handle the diversity and complexity of legal texts. It can retrieve from multiple sources—statutes, judicial rulings, legal doctrines—and present consolidated outputs that reduce the time and effort needed for legal research. In contexts where laws are updated frequently or where legal reasoning requires multi-document understanding, such as in complex case analysis, RAG provides an efficient, scalable alternative to traditional systems.

The potential benefits are particularly evident when applied to under-resourced legal systems or multilingual environments, such as in Algeria, where legal texts exist in both Arabic and French, and digital legal resources remain fragmented. Here, RAG could offer a pathway toward more accessible, interpretable, and comprehensive legal information services\cite{asteraRAG}.

\section{Task and Evaluation}
The rapid development and increasing use of Retrieval-Augmented Generation (RAG) models in NLP have made their evaluation a critical area of research within the LLM community. This section covers the primary downstream tasks associated with RAG, the datasets used, and the methods for evaluating RAG systems \cite{zhou2020trustworthiness}.
\subsection{Factuality}
\begin{itemize}
	\item Fact-Checking Benchmarks: Using datasets like FEVER and SQuAD to evaluate the model's ability to identify and correct factual errors.
	\item Adversarial Testing: Creating adversarial examples to test the model's robustness against misleading information.
	\item Contextual Understanding: Assessing the model's ability to understand the context of a query and provide accurate answers.
\end{itemize}
\subsection{ Robustness}
\begin{itemize}
	\item Noise Injection: Introducing noise into the retrieved documents to test the model's ability to handle imperfect information.
	\item Adversarial Attacks: Evaluating the model's resilience to attacks that aim to manipulate its outputs.
	\item Domain Adaptation: Testing the model's ability to adapt to new domains and data distributions.
\end{itemize}
\subsection{Fairness}
\begin{itemize}

	\item Bias Detection: Identifying and quantifying biases in the training data and model outputs.
	\item Fairness Metrics: Using metrics like demographic parity and equalized odds to evaluate the model's fairness.
	\item Mitigation Techniques: Implementing techniques like debiasing and fairness constraints to mitigate biases.
\end{itemize}
\subsection{Objective Metrics:}
\begin{itemize}
	\item Accuracy: Precision, recall, and F1-score for evaluating the correctness of the generated responses.
	\item Consistency: Measuring the consistency of the model's outputs across different queries and contexts.
	\item Coherence: Assessing the coherence and fluency of the generated text.
\end{itemize}
\subsection{Subjective Metrics}
\begin{itemize}
	\item Human Evaluation: Using human raters to evaluate the quality of the generated responses.
	\item User Studies: Conducting user studies to gather feedback on the user experience
\end{itemize}
\section{ Challenges and Limitations in (RAG)}
While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval} \cite{gupta2024comprehensive}.
\subsection{Scalability and Efficiency}
A major challenge for RAG models lies in their scalability. Since retrieval components depend on external databases, efficiently handling large and constantly expanding datasets demands robust retrieval algorithms. Additionally, the high computational and memory requirements pose difficulties for deploying RAG models in real-time or resource-limited environments
\subsection{Noisy Retrieval Results}
The quality of retrieval results can significantly impact the performance of Retrieval-Augmented Generation (RAG) models. Several factors contribute to noisy retrieval, including the methods used for indexing, the way queries are formulated, and the characteristics of the underlying datasets. When irrelevant or noisy information is retrieved, it may result in the generation of hallucinated or factually inaccurate responses, further compromising the reliability of the system. Additionally, the model’s ability to correctly interpret the retrieved information can be hindered, particularly when the information lacks proper formatting or contains inconsistencies. 
\subsection{Bias and Fairness} 
Like other machine learning models, Retrieval-Augmented Generation (RAG) systems can inherit biases from the datasets they rely on for retrieval. These biases may be amplified during the generation process, potentially leading to biased or harmful outputs. Addressing this issue by developing effective techniques to mitigate bias in both the retrieval and generation phases remains an ongoing challenge.
\subsection{Extra Overhead}
\begin{itemize}
	\item Computational Cost: Retrieval and processing additional information can increase computational costs, especially for large-scale models and complex queries.
	\item Latency: Retrieval and processing can introduce latency, impacting the real-time performance of RAG systems.
	\item Complexity: Implementing and deploying RAG systems requires careful consideration of various factors, including data preparation, model selection, and system architecture.
\end{itemize}

\subsection{Long Context Generation:}

\begin{itemize}
	\item Context Length Limits: LLMs have limitations on the amount of context they can process at once.
	\item Information Loss: Long documents may be truncated or summarized, leading to loss of important information.
	\item Computational Cost: Processing long contexts can be computationally expensive.
\end{itemize}
\newpage
\section{Conclusion}

In this chapter, we explored the core principles and mechanisms of Retrieval-Augmented Generation (RAG), emphasizing its potential to enhance language models through access to external knowledge sources during inference. We examined its general architecture, key stages, and its evolution from naive to more modular frameworks.

A particular focus was placed on the application of RAG in the legal domain, where we highlighted its potential for improving legal information retrieval by enabling more contextualized and semantically rich access to case law, statutes, and scholarly texts. We also outlined the main challenges and limitations facing RAG models, including noisy retrieval, bias, scalability.
